Relative Stable Maps in Genus One via Central Alignments

### 2019-01-28 GenusOneRelativeMaps.tex
fgt on drc, streamlined the description of tropical rays of type Yc0, added recursion procedure from my thesis

\subsection{A desingularisation of punctured log maps}
Describe an analogous desingularisation for punctured log maps. The degree zero case seems more delicate and complicated, even though it should just be a substack of a moduli space of marked curves after all (the log structure is possibly trickier).
\begin{dfn}
 A $(z,g,n)$-punctured log map is a punctured log map where the markings with negative contact order to $H$ have an extra label, either $z$ (the tropical map sends them to $z$ero), or $g$ (stands for gluing), or $n$ (stands for nihil, nothing...)
\end{dfn}
We would like to extend the minimal monoid in order to account for some of the punctures, i.e. take the sum of $p^*\mathcal M_{C^\circ}$ (for every such puncture $p$) fibered over the log structure of the base, which seems to have the same effect as topping those edges with a uertex (which doesn't carry any other information). This looks to me like sprouting a $\PP^1$ at those puncture. Now, the point of the $(z,g,n)$-labelling is that: we only extend the log structure at the $z$ and $g$ punctures; uertices labelled with a $z$ are sent under the tropical map to $0$, so continuity applies to them (therefore the extension of the log structure is fake because that edge length is determined by the position of the adjacent vertex); while uertices labelled with an $n$ are free to roam about. Finally, we perform a radial alignment (and we take rays). This distinction comes about when you split your centrally aligned map according to the strict interior of the $\delta$-circle and the outside; the intersection of a tree with the $\delta$-circle can either be a vertex over $0$, or another vertex (corresponding to a degeneration of the genus zero map outside), or not a vertex.

@@

In our setting, when dealing with truly punctured maps, the associated logarithmic stratum of the target will always be the divisor $H$. Consider a moduli space
\begin{equation*} \Mpunct{g}{\alpha}{H}{d} \end{equation*}
where $g \in \{0,1\}$ and $\alpha$ is a vector of tangency orders, some of which are negative. Here $H$ is given the log structure pulled back from $\PP^N$; since $H$ is not log smooth, the natural obstruction theory on the above moduli space is not necessarily perfect. On the other hand, we will see that this space is birational to the moduli space
\begin{equation*} \ol\Mcal^{\alpha}_{g,n}(H,d) \subseteq \M{g}{n}{H}{d} \end{equation*}
of ordinary stable maps to $H$ satisfying the condition:
\begin{equation*} f^*\OO_H(1) \cong \OO_C(\Sigma_i \alpha_ix_i).\end{equation*}
This is referred to as the \emph{double ramification locus} and has been studied extensively. It is a closed substack of virtual codimension $g$ in $\M{g}{n}{H}{d}$. In the genus zero case it is equal to the entire space
\begin{equation*} \ol\Mcal^{\alpha}_{0,n}(H,d) = \M{0}{n}{H}{d} \end{equation*}
which is smooth. Thus integrals over $\Mpunct{0}{\alpha}{H}{d}$ are the same as integrals over $\M{0}{n}{H}{d}$. The genus one case is more delicate; see \S \ref{} below.

@@

Review here the process of gluing for punctured maps. In particular the basic monoid should be modified in order to make the relevant evaluations log morphisms. Claim: evaluations are strict. Consequence: the fiber product in the category of log stacks is fine. We have then to apply saturation. This is a finite morphism of degree... (I think it could be $\frac{\prod m^{(i)}}{lcm(m^{(i)})}$).

@@

\subsection{Recursive description of the divisors: type $C_0$}
On the other hand, when there is a contracted elliptic subcurve - and it will be contracted into the hyperplane, because otherwise it wouldn't be generic, by density of the nice locus in $\VZ_{1,\alpha^\prime}(\PP^N|H,d^\prime)$ - the picture becomes more complicated due to the alignment. The combs may break (multiple times) and they may point in opposite directions. We label these loci $\mathcal{Y}_c^0$.
\begin{figure}
 \tikz{
 \draw (0,0) circle(2pt)[fill=black] (0,2) circle(2pt)[fill=black] (0,4) circle(2pt)[fill=black] (0,6) circle(2pt)[fill=black] (0,-2) circle(2pt)[fill=black] (1.1,1.78) circle(2pt)[fill=black] (3.42,1.42) circle(2pt)[fill=black] (.58,-1.42) circle(2pt)[fill=black];
 \draw (2,0) -- node[above]{4} (0,0) (2,0) --node[above]{2} (1.1,1.78) --node[above]{2} (0,2) (1.1,1.78) --node[above]{1} (0,4) (2,0) --node[above]{1} (3.42,1.42) (2,0) --node[above]{1} (.58,-1.42) (3.42,1.42) --node[above]{5} (0,6) (.58,-1.42) --node[above]{3} (0,-2);
 \draw[red] (2,0) circle(2);
 \draw (2,0) circle(2pt)[fill=white];
 \draw[->] (0,-3) circle(2pt)[fill=black] edge (4,-3);
 }
\end{figure}
\begin{dfn}
 A degree-aligned genus one tropical map $\phi$ to $\mathbb R_{\geq 0}$ is ... with a circle of radius $\delta$ around the core satisfying:
 \begin{enumerate}
  \item the circle of radius $\delta$ passes through at least one vertex of positive degree, and no vertex of positive degree is contained in its strict interior;
  \item every circle of radius $<\delta$ around the core passes through (either none or) at least two vertices;\footnote{The subdivisions we perform before picking a ray correspond to aligning the log structure, i.e. imposing $\lambda(v)=\lambda(v^\prime)$ for some $v,v^\prime$.}
  \item every edge heading out of the circle of radius $\delta$ goes directly to a vertex in $\phi^{-1}(0)$.
 \end{enumerate}
\end{dfn}

@@

>>>>

%\subsection{A desingularisation of punctured log maps}
%Describe an analogous desingularisation for punctured log maps. The degree zero case seems more delicate and complicated, even though it should just be a substack of a moduli space of marked curves after all (the log structure is possibly trickier).
%\begin{dfn}
% A $(z,g,n)$-punctured log map is a punctured log map where the markings with negative contact order to $H$ have an extra label, either $z$ (the tropical map sends them to $z$ero), or $g$ (stands for gluing), or $n$ (stands for nihil, nothing...)
%\end{dfn}
%We would like to extend the minimal monoid in order to account for some of the punctures, i.e. take the sum of $p^*\mathcal M_{C^\circ}$ (for every such puncture $p$) fibered over the log structure of the base, which seems to have the same effect as topping those edges with a uertex (which doesn't carry any other information). This looks to me like sprouting a $\PP^1$ at those puncture. Now, the point of the $(z,g,n)$-labelling is that: we only extend the log structure at the $z$ and $g$ punctures; uertices labelled with a $z$ are sent under the tropical map to $0$, so continuity applies to them (therefore the extension of the log structure is fake because that edge length is determined by the position of the adjacent vertex); while uertices labelled with an $n$ are free to roam about. Finally, we perform a radial alignment (and we take rays). This distinction comes about when you split your centrally aligned map according to the strict interior of the $\delta$-circle and the outside; the intersection of a tree with the $\delta$-circle can either be a vertex over $0$, or another vertex (corresponding to a degeneration of the genus zero map outside), or not a vertex.

@@

In our setting, when dealing with truly punctured maps, the associated logarithmic stratum of the target will always be the divisor $H$. Consider a moduli space
\begin{equation*} \Mpunct{g}{\alpha}{H}{d} \end{equation*}
where $g \in \{0,1\}$ and $\alpha$ is a vector of tangency orders, some of which are negative. Here $H$ is given the log structure pulled back from $\PP^N$; since $H$ is not log smooth, the natural obstruction theory on the above moduli space is not necessarily perfect. On the other hand, we will see that this space is birational to the moduli space
\begin{equation*} \ol\Mcal^{\alpha}_{g,n}(H,d) \subseteq \M{g}{n}{H}{d} \end{equation*}
of ordinary stable maps to $H$ satisfying the condition:
\begin{equation*} f^*\OO_H(1) \cong \OO_C(\Sigma_i \alpha_ix_i).\end{equation*}
This is referred to as the \emph{double ramification locus} and has been studied extensively. It is a closed substack of virtual codimension $g$ in $\M{g}{n}{H}{d}$. In the genus zero case it is equal to the entire space
\begin{equation*} \ol\Mcal^{\alpha}_{0,n}(H,d) = \M{0}{n}{H}{d} \end{equation*}
which is smooth. Thus integrals over $\Mpunct{0}{\alpha}{H}{d}$ are the same as integrals over $\M{0}{n}{H}{d}$. The genus one case is more delicate; yet, luckily, we are only interested in the main component of the punctured space.

\begin{lem}
 Alignment and factorisation single out the closure of the locus of punctured maps with smooth source curve.
\end{lem}
\begin{proof}
 If it is true, we might be able to prove it via deformation theory. For this it might be useful to notice that there is a morphism $\Mpunct{g}{\alpha}{H}{d}\to \Mpunct{g}{\alpha}{\PP^N}{d}$ (because $H\subseteq\PP^N$ is strict), which is probably a closed immersion, and the loci with smooth source curve are isomorphic under this map. It seems plausible that the main omponent is log smooth over $B\Gm\subseteq[\Aaff^1/\Gm]$ (or the standard log point).
\end{proof}

\begin{lem}
The main component of punctured maps is birational to the main component of the DR locus.
\end{lem}
We probably want the latter statement to hold under imposing alignment and factorisation (keeping the log structure on the source curve, but forgetting that the map is log).

\begin{lem}
 Forgetting a marking $\fgt_i\colon\ol\Mcal_{1,n}^\alpha(H,d)^{\rm{main}} \to\VZ_{1,n-1}(H,d)$ is generically finite of degree $\alpha_i^2$ - except in one case.
\end{lem}
\begin{proof}
 Let us consider the case $d=0$ first. We may assume the source curve is smooth elliptic $E$. The map \[ \phi\colon E\to \Pic^0(E),\qquad x\mapsto\OO_E\left(\alpha_ix+\sum_{j=1,\ldots,\hat i\ldots,n}\alpha_jp_j\right)\]
 is an isogeny of degree $\alpha_i^2$. The locus of $(C,p_1,\ldots,\hat{p_i},\ldots,p_n)$ such that the kernel of $\phi$ contains one of the points $\{p_1,\ldots,\hat{p_i},\ldots,p_n\}$ is itself a double ramification locus inside $\oM_{1,n}$, hence non-generic - with one exception: namely, when $n=2$, $x=p_1$ is always a solution, but the curve lying above such point bubbles off a $\PP^1$. To see that it does not belong to the closure of the nice locus, notice that the rational function trivialising $\OO_C(\alpha_1p_1-\alpha_2p_2)$ should descend to the cusp, thus having a ramification point at the node; yet its ramification profile is determined by Riemann-Hurwitz, and it is entirely supported on $p_1$ and $p_2$.

 For a different proof: notice that we should obtain the class of the main component from the full double ramification cycle by subtracting the boundary class $[D_{1,\emptyset|0,\{1,\ldots,n\}}]$. The latter pushes forward to $0$ under $\fgt_i$, unless $n=2$. Therefore we may apply $\fgt_{i,*}$ to the Hain-Pixton formula:
 \[DR_1(A)=\frac{1}{2}\left(\sum_{i=1}^na_i^2\psi_i-\sum_{\substack{I\subseteq\{1,\ldots,n\}\\|I|\geq 2}}a_I^2[D_{1,I^c|0,I}]-\frac{1}{12}\delta_0\right),\]
 where $a_I=\sum_{i\in I}a_i$ and $\delta_0=\operatorname{glue}_*([\oM_{0,n+2}])$. From $\psi_j=\fgt_i^*\psi_j+[D_{i,j}]$ for $i\neq j$, and the dilaton equation, we see that $\fgt_{i,*}\psi_j=1$ and $\fgt_{i,*}\psi_i=n-1$. On the other hand, the only surviving boundary classes are $[D_{i,j}]$, and they push down to $1$. Hence the formula pushes down to $\frac{1}{2}\left(\sum_{j\neq i}\alpha_j^2+(n-1)\alpha_i^2-\sum_{j\neq i}(\alpha_j+\alpha_i)^2\right)=-\alpha_i\left(\sum_{j\neq i}\alpha_j\right)=\alpha_i^2.$
\end{proof}

@@

\footnote{(Luca) to be made homogeneous with what comes earlier} Review here the process of gluing for punctured maps. In particular the basic monoid should be modified in order to make the relevant evaluations log morphisms. Claim: evaluations are strict. Consequence: the fiber product in the category of log stacks is fine. We have then to apply saturation. This is a finite morphism of degree... (I think it could be $\frac{\prod m^{(i)}}{lcm(m^{(i)})}$).

@@

\subsection{Recursive description of the divisors: type $C_0$}
On the other hand, when there is a contracted elliptic subcurve - and it will be contracted into the hyperplane, because otherwise it wouldn't be generic, by density of the nice locus in $\VZ_{1,\alpha^\prime}(\PP^N|H,d^\prime)$ - the picture becomes more complicated due to the alignment. The combs may break. We label these loci $\mathcal{Y}_C^0$.
\begin{figure}

 \tikz{
 \draw[red] (2,0) circle(2); %circle
 \draw (0,0) circle(2pt)[fill=black] (0,2) circle(2pt)[fill=black] (0,4) circle(2pt)[fill=black] (0,6) circle(2pt)[fill=black] (0,-2) circle(2pt)[fill=black] (1.1,1.78) circle(2pt)[fill=black] (3.42,1.42) circle(2pt)[fill=black] (.58,-1.42) circle(2pt)[fill=black] (0,1) circle(2pt)[fill=black]; %black vertices
 \draw (2,0) -- node[above]{4} (0,0) (2,0) --node[above]{2} (1.1,1.78) --node[above]{1} (0,2) (1.1,1.78) --node[above]{1} (0,4) (2,0) --node[above]{1} (3.42,1.42) (2,0) --node[above]{3} (.58,-1.42) (3.42,1.42) --node[above]{5} (0,6) (.58,-1.42) --node[above]{1} (0,-2) (2,0) -- node[above]{2} (0,1); %edges
 \draw (.58,-1.42) -- (5,-1.42) node[right]{1} (3.42,1.42) -- (5,1.42) node[right]{2} (2,0) -- (5,.5) node[right]{3} (2,0) -- (5,-.5) node[right]{7}; %markings
 \draw (2,0) circle(2pt)[fill=white]; %core
 \draw[->] (0,-3) circle(2pt)[fill=black] edge (5,-3); %trop(P^N|H)
 }

 \caption{The degrees can be figured out from the balancing equation.}
\end{figure}

In the following we describe the rays of the tropical moduli space.
\begin{lem}
 A degree-aligned genus one tropical map $\phi$ to $\mathbb R_{\geq 0}$ is ... with a circle of radius $\delta$ around the core satisfying:
 \begin{enumerate}
 \item the circle of radius $\delta$ passes through at least one vertex of $\phi^{-1}(0)$ - which necessarily has positive degree - call $m$ its contact order with $H$;
 \item teeth may break only when they intersect the circle of radius $\delta$; in particular, $\circ$ is the only vertex contained in its strict interior, and every edge heading out from the circle goes directly to a vertex of $\phi^{-1}(0)$;
 \item every tooth that starts with contact order $m$ goes directly to a vertex of $\operatorname{circle}(\circ,\delta)\cap\phi^{-1}(0)$, and every other tooth starts with contact order $<m$ (possibly negative).
 \end{enumerate}
\end{lem}
\begin{proof}
 Otherwise there would be more than one parameters.
\end{proof}

@@

 Assume \footnote{(Luca) copied and pasted from my thesis} we want to compute $\langle\tau_{h_1}H^{k_1},\ldots,\tau_{h_1}H^{k_1}\rangle^{\PP^{N-1},\rm{red}}_{1,n,d}$ by writing it as $\frac{1}{d^2}(\psi_1^{\{n+1\}})^{h_1}H^{k_1}\cdots(\psi_n^{\{n+1\}})^{h_n}H^{k_n}\cdot[\VZ{(0,\ldots,0,d)}{\PP^N|H}{d}^\thicksim]$. Let me start from the recursion step $H_1\cdot[\VZ{(0,\ldots,0,d)}{\PP^N|H}{d}]=\ldots:$

 \begin{description}
  \item[$\mathcal Y_a$:] $d\ {\tikz[baseline=-3pt,scale=.7]{
\draw (2,0) node[above]{$\{x_1,x_{n+1}\}$} -- (0,0)node[below]{$d$};
\draw (2,0) circle(2pt)[fill=black];
\draw (0,0) circle(2pt)[fill=white];
\draw (-1,1.5) -- (1,1.5) -- (1,-1.5) -- (-1,-1.5) -- (-1,1.5);
\draw (1,1.5) -- (3,2) -- (3,-1) -- (1,-1.5);
}}$ contributes $\frac{1}{d}\psi_1^{h_1}H^{k_1}\cdots\psi_n^{h_n}H^{k_n}\cdot[\VZ{(d,\ldots,0)}{\PP^N|H}{d}]$; if this is the curve configuration, while there is any other marking at level one, then $\fgt_{n+1}$ has positive dimensional fibers; otherwise there must be another component of positive horizontal degree (these contributions are computed inductively on $d$);
\item[$\mathcal Y_b$:] is an entirely rational story;
\item[$\mathcal Y_c$:] {\tikz[baseline=-3pt,scale=.7]{
\draw (2,0) node[above]{$\{x_1,\ldots,x_{n+1}\}$} node[below]{$d$};
\draw (2,0) circle(2pt)[fill=white];
\draw (-1,1.5) -- (1,1.5) -- (1,-1.5) -- (-1,-1.5) -- (-1,1.5);
\draw (1,1.5) -- (3,2) -- (3,-1) -- (1,-1.5);
}} is the term we are meant to compute, while all other terms will include a DRC integral of lower degree.
 \end{description}
I shall look one step further into the recursion for computing $[\VZ{(0,\ldots,0,d)}{\PP^N|H}{d}]$: in order to do so consider the formula for $((d-1)\psi_{n+1}+H_{n+1})\cdot[\VZ{(0,\ldots,0,d-1,1)}{\PP^N|H}{d}]$. This produces:
\begin{description}
  \item[$\mathcal Y_a$:] $d\ {\tikz[baseline=-3pt,scale=.7]{
\draw (2,0) node[above]{$\{x_{n+1},x_{n+2}\}$} -- (0,0)node[below]{$d$};
\draw (2,0) circle(2pt)[fill=black];
\draw (0,0) circle(2pt)[fill=white];
\draw (-1,1.5) -- (1,1.5) -- (1,-1.5) -- (-1,-1.5) -- (-1,1.5);
\draw (1,1.5) -- (3,2) -- (3,-1) -- (1,-1.5);
}}$ this is the one we are after; while
$(d-1){\tikz[baseline=-3pt,scale=.7]{
\draw (2,.5) node[above]{$\{x_{k},x_{n+1}\}$} -- (0,0)node[below]{$d$};
\draw (2,-.5) node[above]{$\{x_{n+2}\}$} -- (0,0);
\draw (2,.5) circle(2pt)[fill=black];
\draw (2,-.5) circle(2pt)[fill=black];
\draw (0,0) circle(2pt)[fill=white];
\draw (-1,1.5) -- (1,1.5) -- (1,-1.5) -- (-1,-1.5) -- (-1,1.5);
\draw (1,1.5) -- (3,2) -- (3,-1) -- (1,-1.5);
}}$ gives a lower tangency condition $[\VZ{(0,\ldots,d-1,\ldots,0)}{\PP^N|H}{d}]$; and all other terms have lower degree.
\item[$\mathcal Y_c$:] {\tikz[baseline=-3pt,scale=.7]{
\draw (2,0) node[above]{$\{x_1,\ldots,x_{n+1},x_{n+2}\}$} node[below]{$d$};
\draw (2,0) circle(2pt)[fill=white];
\draw (-1,1.5) -- (1,1.5) -- (1,-1.5) -- (-1,-1.5) -- (-1,1.5);
\draw (1,1.5) -- (3,2) -- (3,-1) -- (1,-1.5);
}} has positive dimensional fibers for $\fgt_{n+1,n+2}$, and all other terms have lower degree.
 \end{description}
 Finally, let me look one step further into the reduction of $[\VZ{(d,\ldots,0)}{\PP^N|H}{d}]$: we have $((d-1)\psi_1^{\{n+1\}}+H_1)\cdot[\VZ{(d-1,\ldots,0,1)}{\PP^N|H}{d}]=\ldots:$
 \begin{description}
  \item[$\mathcal Y_a$:] $d\ {\tikz[baseline=-3pt,scale=.7]{
\draw (2,0) node[above]{$\{x_{1},x_{n+1}\}$} -- (0,0)node[below]{$d$};
\draw (2,0) circle(2pt)[fill=black];
\draw (0,0) circle(2pt)[fill=white];
\draw (-1,1.5) -- (1,1.5) -- (1,-1.5) -- (-1,-1.5) -- (-1,1.5);
\draw (1,1.5) -- (3,2) -- (3,-1) -- (1,-1.5);
}}$ is the one we care about;
$(d-1)\ {\tikz[baseline=-3pt,scale=.7]{
\draw (2,.5) node[above]{$\{x_{k},x_{1}\}$} -- (0,0)node[below]{$d$};
\draw (2,-.5) node[above]{$\{x_{n+1}\}$} -- (0,0);
\draw (2,.5) circle(2pt)[fill=black];
\draw (2,-.5) circle(2pt)[fill=black];
\draw (0,0) circle(2pt)[fill=white];
\draw (-1,1.5) -- (1,1.5) -- (1,-1.5) -- (-1,-1.5) -- (-1,1.5);
\draw (1,1.5) -- (3,2) -- (3,-1) -- (1,-1.5);
}}$ has a lower number of markings (indeed $\psi_1^{\{n+1\}}$ and $\psi_k^{\{n+1\}}$ restrict to $0$ on this locus, and $\ev_1\equiv\ev_k$); while all other terms have lower degree.
\item[$\mathcal Y_c$:] {\tikz[baseline=-3pt,scale=.7]{
\draw (2,0) node[above]{$\{x_1,\ldots,x_{n+1}\}$} node[below]{$d$};
\draw (2,0) circle(2pt)[fill=white];
\draw (-1,1.5) -- (1,1.5) -- (1,-1.5) -- (-1,-1.5) -- (-1,1.5);
\draw (1,1.5) -- (3,2) -- (3,-1) -- (1,-1.5);
}} contributes $\frac{1}{d}\psi_1^{h_1}H^{k_1}\cdots\psi_n^{h_n}H^{k_n}\cdot[\VZ{(0,\ldots,0,d)}{\PP^N|H}{d}^\thicksim]$ so \emph{there is no cancellation}, while all other terms have lower degree.
 \end{description}
 To sum up, it is possible to determine the reduced genus one restricted invariants of $H$ in terms of those of $\PP^N$ and the genus zero theory, by induction on the degree, number of markings, and total tangency order.
